---
title: "Lecture_6"
author: "Chen, Yen-Ting"
date: "2021/11/2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Metapopulation lab exercise

```{r}
pacman::p_load(dplyr, tidyr, ggplot2, deSolve)
```

This exercise kind of persuades me that this model can actually solve real problems.

$P_{00}$: no plant, untrained soil
$P_{A0}$: plant alive, untrained soil
$P_{AA}$: plant alive, trained soil
$P_{0A}$: plant dead, trained soil

note that p00 + pA0 + pAA + p0A = P = 1 and p00_dt + pA0_dt + pAA_dt + p0A_dt = 0

Probably can test how areas of ecological function changes through time.

Reading assignments only.

```{r}
# ini
ini <- c(p00 = 0.25, pA0 = 0.25, pAA = 0.25, p0A = 0.25)

# flow parameters
parms <- c(dA = 0.1, # soil decay
           RA = 0.8, # plant rep.
           mA = 0.1, # plant mortality 
           CA = 0.5, # soil entrainment
           alpha = 0.7) # RA modifier

time <- seq(0, 300, 0.1)

# patch occupancy
# the pAA + p0A term is the fecundity of alive plants
po_model <- function(time, ini, parms){
  with(as.list(c(parms, ini)), {
    p00_dt <- dA * p0A + mA * pA0 - RA * (pAA + p0A) * p00
    pA0_dt <- RA * (pAA + p0A) * p00 - CA * pA0 - mA * pA0
    pAA_dt <- CA * pA0 - mA * pAA + RA * (pAA + p0A) * alpha * p0A
    p0A_dt <- - dA * p0A + mA * pAA - RA * (pAA + p0A) * alpha * p0A

    return(list(c(p00_dt, pA0_dt, pAA_dt, p0A_dt)))
  })
}


output <- 
  ode(func = po_model,
      times = time,
      parms = parms,
      y = ini) %>% 
  data.frame

output_long <- 
  output %>% 
  pivot_longer(cols = 2:5, 
               names_to = "state", 
               values_to = "values") %>% 
  mutate(state = factor(state, c("pAA", "pA0", "p0A", "p00")))




head(output)
```


```{r}
output_long %>% 
  ggplot() +
  geom_area(aes(x = time, y = values, fill = state))

output_long %>% 
  ggplot() +
  geom_path(aes(x = time, y = values, color = state)) +
  ylim(c(0,1))
```
