---
title: "Exponential growth"
author: "Chen, Yen-Ting"
date: "2021/10/5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#

```{r}
pacman::p_load(deSolve,
               tidyr,
               dplyr,
               ggplot2)
```

To analyze models, analytical solutions or numerical solutions could be done.

Analytical solutions can be sought with simply pen and paper; however, with the terms piles up, it would take years to reach the solution.
Luckily, with technological advancements, numerical solutions can be achieved with computational power. 


desolve

1. model specification

During this stage, 

create a function via specific 
- times (time duration)
- state (initial conditions)
- parameters (parameters to be included)

2. model application
the function `ode` needs 4 argurements for it to work.

```{r}
# 1. specification

exponential_model <- function(times, state, parms){
  with(as.list(c(state, parms)), {
    dN_dt <- r*N # exponential growth equation
    return(list(c( dN_dt)))
  })
}

# 2. application

# set inputs
times <- seq(0, 10, 0.1)
state <- c(N = 10)
parms <- c(r = 1.5)

pop_size<-
  ode(func = exponential_model, # remember not to include the () after the model
      times = times,
      y = state, 
      parms = parms)
class(pop_size)
```

```{r}
# visualization

ggplot(as.data.frame(pop_size), aes(x = time, y = N)) +
  geom_point()

ggplot(as.data.frame(pop_size), aes(x = time, y = N)) +
  geom_point() +
  scale_x_continuous(limits = c(0, 10.5), expand = c(0, 0)) + # expand = for removing the padding around the main graph
  scale_y_log10(breaks = scales::trans_breaks("log10", function(x) 10^x)(c(10, 10^7)),
                labels = scales::trans_format("log10", scales::math_format(10^.x)),
                expand = c(0,0))

```

